---
title: "AMR Data Visualization in R"
subtitle: "A Step-by-Step Teaching Guide for AMR Surveillance Professionals"
author: "Alfred Bortey (DR. AMR)"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    number_sections: true
    theme: flatly
    highlight: tango
    df_print: paged
    code-tools: true
    #code_folding: show
---

```{css, echo=FALSE}
/* ── custom teaching styles ────────────────────────────────────── */
body { font-family: 'Segoe UI', Helvetica, sans-serif; font-size: 15px; color: #2c3e50; }
h1, h2, h3 { color: #1A4A8A; }
h1 { border-bottom: 3px solid #E8B800; padding-bottom: 8px; }

.learn-box {
  background: #EBF3FB; border-left: 5px solid #1A4A8A;
  padding: 12px 16px; border-radius: 4px; margin: 16px 0;
}
.warn-box {
  background: #FFF8E1; border-left: 5px solid #E8B800;
  padding: 12px 16px; border-radius: 4px; margin: 16px 0;
}
.tip-box {
  background: #EAFAF1; border-left: 5px solid #27AE60;
  padding: 12px 16px; border-radius: 4px; margin: 16px 0;
}
.danger-box {
  background: #FDEDEC; border-left: 5px solid #C0392B;
  padding: 12px 16px; border-radius: 4px; margin: 16px 0;
}
.exercise-box {
  background: #F5EEF8; border: 2px solid #8E44AD;
  padding: 12px 16px; border-radius: 6px; margin: 16px 0;
}
pre { background: #1e2330; color:  #1A4A8A; border-radius: 6px; padding: 14px; }
code { color: #8B0000; background: #f8f2f2; padding: 1px 4px; border-radius: 3px; }
pre code { color: #1e2339; background: transparent; }
.caption { font-size: 13px; color: #7f8c8d; font-style: italic; text-align: center; margin-top: -10px; }
table { border-collapse: collapse; width: 100%; font-size: 14px; }
th { background: #1A4A8A; color: white; padding: 8px 12px; }
td { padding: 6px 12px; border-bottom: 1px solid #e8e8e8; }
tr:nth-child(even) { background: #f4f8fd; }
```

---

::: {.learn-box}
**Welcome to AMR Data Visualization**

By the end of this guide you will be able to:

- Import real-world AMR data from Excel into R (including messy, unclean data)
- Clean and prepare the data for analysis
- Create five core chart types used in AMR surveillance
- Interpret each chart in a public health context
- Save publication-quality figures



**Time needed:** Approximately 3–4 hours (work through it section by section).
:::

---

# Before You Start

## What you need installed

Run this once to install all packages used in this guide.  
You only need to do this **one time** — after that, just load the packages with `library()`.

```{r install, eval=FALSE}
# ── Run this block ONCE to install all required packages ──────────────────
# (Remove the # from the install.packages line below if needed)

# install.packages(c(
#   "readxl",      # read Excel files
#   "dplyr",       # data manipulation
#   "tidyr",       # reshaping data
#   "stringr",     # text/string cleaning
#   "lubridate",   # working with dates
#   "ggplot2",     # all charts
#   "scales",      # percentage labels on axes
#   "viridis",     # colourblind-friendly palettes
#   "patchwork",   # combine multiple plots
#   "janitor",     # clean column names
#   "knitr",       # table formatting in reports
#   "kableExtra"   # extra table styling
# ))
```

## Load packages

Every time you open R and start working, you need to load the packages:

```{r setup, message=FALSE, warning=FALSE}
library(readxl)      # reading Excel files
library(dplyr)       # data manipulation: filter, mutate, group_by, summarise
library(tidyr)       # reshaping: pivot_longer, pivot_wider
library(stringr)     # string cleaning: str_to_upper, str_trim, str_detect
library(lubridate)   # date parsing: dmy(), mdy(), ymd()
library(ggplot2)     # all visualizations
library(scales)      # axis label formatting (%, commas)
library(viridis)     # colour palettes accessible to colourblind readers
library(patchwork)   # combine multiple ggplot2 charts side by side
library(janitor)     # clean_names() standardises column names
library(knitr)       # kable() for neat tables
library(kableExtra)  # extra kable formatting
```

## Set a consistent plot theme

We define one theme here and reuse it on every chart so they all look consistent.
Think of this like setting a style template in PowerPoint — once, not on every slide.

```{r theme}
# ── Consistent chart theme ────────────────────────────────────────────────
amr_theme <- theme_minimal(base_size = 13) +
  theme(
    plot.title    = element_text(colour = "#1A4A8A", face = "bold",
                                 size = 14, hjust = 0),
    plot.subtitle = element_text(colour = "#7f8c8d", size = 11, hjust = 0),
    plot.caption  = element_text(colour = "#aaaaaa", size = 9,
                                 hjust = 1, face = "italic"),
    axis.title    = element_text(face = "bold", size = 11),
    axis.text     = element_text(size = 10),
    legend.position  = "bottom",
    legend.title     = element_text(face = "bold", size = 10),
    panel.grid.minor = element_blank(),
    strip.text       = element_text(face = "bold", colour = "#1A4A8A")
  )

# Apply this theme globally — all charts will use it automatically
theme_set(amr_theme)
```

---

# Importing Data from Excel

::: {.warn-box}
**Why this step matters**

In real AMR surveillance, data rarely arrives in a perfectly clean format. It typically comes from:

- A WHONET export (`.csv` or `.dbf`)
- A lab information system (LIS) export (`.xlsx`)
- A paper form that someone typed into Excel
- A DHIMS2 or HMIS download

Our teaching Excel file (`AMR_Teaching_Data.xlsx`) has been designed to look like real lab data — with inconsistent names, mixed date formats, blank rows, and extra columns. We will import it and clean it together.
:::

## Looking at the Excel workbook first

Our workbook has **five sheets**. Let's see what is available:

```{r see-sheets}
# ── Path to your Excel file ───────────────────────────────────────────────
# IMPORTANT: Change this path to wherever you saved the Excel file.
# Examples:
#   Windows:  "C:/Users/YourName/Documents/AMR_Teaching_Data.xlsx"
#   Mac/Linux: "/home/yourname/AMR_Teaching_Data.xlsx"

excel_file <- "AMR_Teaching_Data.xlsx"

# List all sheets in the workbook
excel_sheets(excel_file)
```

The output shows all available sheets. We will use:

| Sheet | What it contains |
|-------|-----------------|
| `Raw_AST_Data` | 130 isolate records — messy, realistic |
| `Consumption` | Monthly antibiotic use (DDD & DOT) |
| `MIC_Values` | Raw MIC values for histogram and box plot |
| `Quarterly_R` | Pre-aggregated quarterly resistance rates |

## Import Sheet 1: Raw AST data (the messy sheet)

```{r import-raw}
# ── Import the raw AST data ───────────────────────────────────────────────
# skip = 0 means start from row 1 (the header row)
# .name_repair = "unique" handles any duplicate column names

raw_ast <- read_excel(
  excel_file,
  sheet = "Raw_AST_Data",
  skip  = 0,
  .name_repair = "unique"
)

# Quick look at the data
cat("Rows:", nrow(raw_ast), "\n")
cat("Columns:", ncol(raw_ast), "\n\n")

# Show first 8 rows
head(raw_ast, 8)
```

::: {.warn-box}
**What you will notice in this data (and why it is realistic)**

- Organism names are inconsistent: `"Klebsiella pneumoniae"`, `"K. pneumoniae"`, `"KLEBSIELLA PNEUMONIAE"` all appear
- Some AST results say `"R"`, others say `"r"` or `"res"` — same meaning, different format
- Dates are in mixed formats: `2022-01-05`, `05/01/2022`, `Jan 05 2022`
- There are blank rows (common in Excel exports)
- There is a row that is a repeated header (another common Excel artifact)
- Columns like `EXTRA_COL_IGNORE` and `Notes` are irrelevant

This is normal. The next section shows you how to fix all of this.
:::

---

# Data Cleaning: Making the Data Ready for Analysis

Cleaning data is not glamorous, but it is **the most important step**. Bad data produces
misleading results. This section walks through each cleaning step and explains *why* each
one is necessary.

## Step 1 — Remove completely blank rows

Blank rows cause errors later when R tries to count isolates or calculate percentages.

```{r clean-blanks}
# ── Remove rows where ALL key columns are empty ───────────────────────────
raw_ast <- raw_ast %>%
  # filter() keeps rows where the condition is TRUE
  # !is.na() means "is NOT missing"
  # We check PatientID — if it is missing, the row is blank or a repeat header
  filter(!is.na(PatientID) & PatientID != "PatientID")
  # PatientID != "PatientID" removes any rows where the header was repeated

cat("Rows remaining after removing blanks:", nrow(raw_ast), "\n")
```

## Step 2 — Standardise column names

Column names with spaces, capital letters or special characters cause problems in R code.
`janitor::clean_names()` converts everything to lowercase with underscores:

```{r clean-names}
raw_ast <- raw_ast %>%
  clean_names()

# See the new column names
names(raw_ast)
head(raw_ast)
```

Notice how `CollectionDate` became `collection_date`, `PatientID` became `patient_id`, etc.
This makes typing easier and prevents errors.

## Step 3 — Remove irrelevant columns

```{r clean-cols}
# Drop columns we do not need
raw_ast <- raw_ast %>%
  dplyr::select(-extra_col_ignore, -notes)
  # select(-colname) removes a column
```

## Step 4 — Standardise organism names

This is one of the most important steps in AMR surveillance data cleaning.
We group all spelling variants into one standard name.

```{r clean-organisms}
# ── Standardise organism names ────────────────────────────────────────────
# str_detect() searches for a pattern inside a text string
# str_to_upper() converts text to ALL CAPS so we can match regardless of case

raw_ast <- raw_ast %>%
  mutate(
    # First, convert everything to uppercase for matching
    org_upper = str_to_upper(str_trim(organism)),

    # Then assign a standard name based on what we detect
    organism_std = case_when(
      str_detect(org_upper, "KLEBSIELLA|K\\. PNEUM")   ~ "Klebsiella pneumoniae",
      str_detect(org_upper, "ESCHERICHIA|E\\. COLI|E\\.COLI") ~ "Escherichia coli",
      str_detect(org_upper, "STAPHYLOCOCCUS|STAPH|MRSA|S\\. AUREUS") ~ "Staphylococcus aureus",
      str_detect(org_upper, "PSEUDOMONAS|P\\. AERUG") ~ "Pseudomonas aeruginosa",
      str_detect(org_upper, "ACINETOBACTER|A\\. BAUM") ~ "Acinetobacter baumannii",
      TRUE ~ organism   # keep original if nothing matched (so we can review)
    )
  ) %>%
  dplyr::select(-org_upper)  # remove the temporary column

# Check: how many unique organism names do we have now?
cat("Unique organism names after standardisation:\n")
print(table(raw_ast$organism_std))
```

## Step 5 — Standardise AST results (R/I/S)

```{r clean-ris}
# ── Standardise AST result notation ──────────────────────────────────────
# All columns from 'amp' to 'col' contain AST results
# We want only R, I, S, or NA

ast_cols <- c("amp","amc","ctx","mem","cip","gen","sxt","van","oxa","col")

standardise_ris <- function(x) {
  # Convert to uppercase and trim whitespace first
  x <- str_to_upper(str_trim(as.character(x)))
  case_when(
    x %in% c("R", "RES", "RESISTANT")       ~ "R",
    x %in% c("I", "INT", "INTERMEDIATE")    ~ "I",
    x %in% c("S", "SUS", "SUSCEPTIBLE")     ~ "S",
    x %in% c("NA", "N/A", "", "NULL")       ~ NA_character_,
    TRUE                                     ~ NA_character_  # anything else → NA
  )
}

raw_ast <- raw_ast %>%
  mutate(across(
    all_of(ast_cols[ast_cols %in% names(raw_ast)]),
    standardise_ris
  ))

# Verify: should only see R, I, S, or NA now
cat("Unique CIP values after standardisation:\n")
print(table(raw_ast$cip, useNA = "ifany"))
```

## Step 6 — Parse dates

Mixed date formats are extremely common in lab data across Africa and LMICs.
`lubridate` can handle multiple formats with `parse_date_time()`:

```{r clean-dates}
# ── Parse dates from multiple formats ────────────────────────────────────
# parse_date_time() tries each format in order until one works
raw_ast <- raw_ast %>%
  mutate(
    collection_date = parse_date_time(
      collection_date,
      orders = c("ymd", "dmy", "mdy", "d b Y", "d-b-Y"),
      quiet  = TRUE  # suppress warnings for each try
    ),
    # Create time period columns for trend analysis
    year    = year(collection_date),
    quarter = paste0("Q", quarter(collection_date), "-", year(collection_date)),
    month   = month(collection_date, label = TRUE)
  )

cat("Date range in dataset:\n")
cat("Earliest:", as.character(min(raw_ast$collection_date, na.rm=TRUE)), "\n")
cat("Latest:  ", as.character(max(raw_ast$collection_date, na.rm=TRUE)), "\n")
```

## Step 7 — Apply the First Isolate Rule

::: {.learn-box}
**What is the First Isolate Rule?**

When the same patient has the same organism cultured multiple times (e.g., an ICU
patient with a persistent K. pneumoniae bloodstream infection cultured every week),
including all of those results in resistance calculations **inflates** the resistance rate.

The **CLSI M39-A4 guideline** requires that for cumulative susceptibility reporting
(antibiograms), you keep only the **first isolate per patient per organism** within
each analysis period.
:::

```{r first-isolate}
# ── Apply First Isolate Rule ──────────────────────────────────────────────
cat("Records BEFORE First Isolate Rule:", nrow(raw_ast), "\n")

clean_data <- raw_ast %>%
  arrange(patient_id, organism_std, specimen_type, collection_date) %>%
  group_by(patient_id, organism_std) %>%
  mutate(
    isolate_rank  = row_number(),    # 1 = first isolate, 2 = second, etc.
    is_first      = (isolate_rank == 1)
  ) %>%
  ungroup() %>%
  filter(is_first) %>%              # keep only first isolates
  dplyr::select(-isolate_rank, -is_first)

cat("Records AFTER First Isolate Rule:", nrow(clean_data), "\n")
cat("Isolates excluded:", nrow(raw_ast) - nrow(clean_data), "\n")
```

# Diiferent code for first isolate rule
This considers situations where you might have genuinely different isolates from a patient because they have multiple infections from different sites or sources.

```{r first-isolate-Different-source}
# ── Apply First Isolate Rule ──────────────────────────────────────────────
cat("Records BEFORE First Isolate Rule:", nrow(raw_ast), "\n")

clean_data <- raw_ast %>%
  arrange(patient_id, organism_std, specimen_type, collection_date) %>%
  group_by(patient_id, organism_std, specimen_type) %>%
  mutate(
    isolate_rank  = row_number(),    # 1 = first isolate, 2 = second, etc.
    is_first      = (isolate_rank == 1)
  ) %>%
  ungroup() %>%
  filter(is_first) %>%              # keep only first isolates
  dplyr::select(-isolate_rank, -is_first)

cat("Records AFTER First Isolate Rule:", nrow(clean_data), "\n")
cat("Isolates excluded:", nrow(raw_ast) - nrow(clean_data), "\n")
```

## Summary of cleaned data

Let's check what we have:

```{r clean-summary}
# Organism frequency table
organism_summary <- clean_data %>%
  count(organism_std, name = "n_isolates") %>%
  mutate(percent = round(n_isolates / sum(n_isolates) * 100, 1)) %>%
  arrange(desc(n_isolates))

kable(organism_summary,
      col.names = c("Organism", "N isolates", "%"),
      caption   = "Table 1. Organism frequency after cleaning and deduplication") %>%
  kable_styling(bootstrap_options = c("striped","hover","condensed"),
                full_width = FALSE)
```

---

# Visualization 1: Bar Plot — Comparing Resistance Rates

::: {.learn-box}
**When to use a bar plot in AMR surveillance**

Bar plots are used when you want to **compare a single number** (like % resistant)
across different categories (like organisms, wards, or antibiotics).

They are the most common chart in AMR reports because they are easy to read and
communicate clearly to clinical and non-technical audiences.

**AMR examples:**
- Ciprofloxacin resistance in K. pneumoniae vs E. coli vs P. aeruginosa
- MRSA rates across different wards
- Resistance to different antibiotics for the same organism
:::

## Calculate resistance rates

Before plotting, we need to calculate what we want to show.
We will calculate **% resistant** with a 95% confidence interval (CI).

```{r bar-calc}
# ── Calculate resistance rates with 95% Wilson Score CI ──────────────────
# Why Wilson Score CI? It works correctly even when n is small or
# the proportion is near 0% or 100% — both very common in AMR data.

calc_resistance_ci <- function(data, organism_name, antibiotic) {
  col <- antibiotic   # the column in our data

  data %>%
    filter(organism_std == organism_name, !is.na(.data[[col]])) %>%
    summarise(
      organism   = organism_name,
      antibiotic = toupper(antibiotic),
      n_tested   = n(),
      n_R        = sum(.data[[col]] == "R", na.rm = TRUE),
      .groups    = "drop"
    ) %>%
    mutate(
      pct_R = round(n_R / n_tested * 100, 1),
      # Wilson Score CI calculation
      p      = n_R / n_tested,
      z      = 1.96,
      denom  = 1 + z^2 / n_tested,
      centre = (p + z^2 / (2 * n_tested)) / denom,
      margin = z * sqrt(p*(1-p)/n_tested + z^2/(4*n_tested^2)) / denom,
      ci_lo  = round(pmax(0, (centre - margin)*100), 1),
      ci_hi  = round(pmin(100, (centre + margin)*100), 1)
    ) %>%
    dplyr::select(organism, antibiotic, n_tested, n_R, pct_R, ci_lo, ci_hi)
}

# Calculate for ciprofloxacin across all priority organisms
cip_resistance <- bind_rows(
  calc_resistance_ci(clean_data, "Klebsiella pneumoniae",  "cip"),
  calc_resistance_ci(clean_data, "Escherichia coli",       "cip"),
  calc_resistance_ci(clean_data, "Staphylococcus aureus",  "cip"),
  calc_resistance_ci(clean_data, "Pseudomonas aeruginosa", "cip"),
  calc_resistance_ci(clean_data, "Acinetobacter baumannii","cip")
)

# Show the table
kable(cip_resistance,
      col.names = c("Organism","Antibiotic","N tested","N resistant",
                    "% Resistant","CI lower","CI upper"),
      caption   = "Table 2. Ciprofloxacin resistance rates with 95% Wilson Score CI") %>%
  kable_styling(bootstrap_options = c("striped","hover","condensed"))
```

#STREAMLINED CALCULATION
Without having specify ncalculation for each antibiotic like in the previous section.

```{r bar-calc-multiple}
# ── Calculate resistance rates with 95% Wilson Score CI ──────────────────
# Why Wilson Score CI? It works correctly even when n is small or
# the proportion is near 0% or 100% — both very common in AMR data.

calc_resistance_ci <- function(data, organism_name, antibiotic) {
  col <- antibiotic   # the column in our data

  data %>%
    filter(organism_std == organism_name, !is.na(.data[[col]])) %>%
    summarise(
      organism   = organism_name,
      antibiotic = toupper(antibiotic),
      n_tested   = n(),
      n_R        = sum(.data[[col]] == "R", na.rm = TRUE),
      .groups    = "drop"
    ) %>%
    mutate(
      pct_R = round(n_R / n_tested * 100, 1),
      # Wilson Score CI calculation
      p      = n_R / n_tested,
      z      = 1.96,
      denom  = 1 + z^2 / n_tested,
      centre = (p + z^2 / (2 * n_tested)) / denom,
      margin = z * sqrt(p*(1-p)/n_tested + z^2/(4*n_tested^2)) / denom,
      ci_lo  = round(pmax(0, (centre - margin)*100), 1),
      ci_hi  = round(pmin(100, (centre + margin)*100), 1)
    ) %>%
    dplyr::select(organism, antibiotic, n_tested, n_R, pct_R, ci_lo, ci_hi)
}

#Select antibiotic columns
antibiotics <- names(clean_data)[
  names(clean_data) %in% c("cip","gen","amk","ctx","fox","mem")
]
# Priority organisms
priority_organisms <- c(
  "Klebsiella pneumoniae",
  "Escherichia coli",
  "Staphylococcus aureus",
  "Pseudomonas aeruginosa",
  "Acinetobacter baumannii"
)

# Create combinations
all_combinations <- expand.grid(
  organism = priority_organisms,
  antibiotic = antibiotics,
  stringsAsFactors = FALSE
)

# Calculate resistance
all_resistance <- all_combinations %>%
  rowwise() %>%
  do(
    calc_resistance_ci(
      clean_data,
      .$organism,
      .$antibiotic
    )
  ) %>%
  ungroup()

# Show the table
kable(all_resistance,
      col.names = c("Organism","Antibiotic","N tested","N resistant",
                    "% Resistant","CI lower","CI upper"),
      caption = "Table 2. Resistance rates with 95% Wilson CI") %>%
  kable_styling(bootstrap_options = c("striped","hover","condensed"))

```
::: {.tip-box}
**Reading the confidence interval**

The 95% CI tells you the range within which the true resistance rate probably lies.
A wide CI (e.g., 20–60%) means you have too few isolates tested — interpret with caution.
A narrow CI (e.g., 44–52%) means you have enough data to be confident in the estimate.

**Rule of thumb:** Only report a resistance rate formally if n ≥ 30 isolates.
:::

## Create the bar plot

```{r bar-plot, fig.width=9, fig.height=5.5}
# ── Add a colour variable based on resistance level ───────────────────────
cip_resistance <- cip_resistance %>%
  mutate(
    res_category = case_when(
      pct_R >= 70 ~ "High (≥70%)",
      pct_R >= 40 ~ "Moderate (40–69%)",
      TRUE        ~ "Low (<40%)"
    ),
    # Short organism labels for the axis
    org_short = case_when(
      organism == "Klebsiella pneumoniae"   ~ "K. pneumoniae",
      organism == "Escherichia coli"        ~ "E. coli",
      organism == "Staphylococcus aureus"   ~ "S. aureus",
      organism == "Pseudomonas aeruginosa"  ~ "P. aeruginosa",
      organism == "Acinetobacter baumannii" ~ "A. baumannii",
      TRUE ~ organism
    )
  )

# ── Build the bar chart ───────────────────────────────────────────────────
p_bar <- ggplot(
  cip_resistance,
  aes(x    = reorder(org_short, -pct_R),  # reorder: tallest bar first
      y    = pct_R,
      fill = res_category)
) +
  # The bars
  geom_col(width = 0.6, colour = "white") +

  # Error bars showing the 95% CI
  geom_errorbar(
    aes(ymin = ci_lo, ymax = ci_hi),
    width = 0.2, colour = "grey30", linewidth = 0.7
  ) +

  # Labels above each bar: percentage and n
  geom_text(
    aes(label = paste0(pct_R, "%\n(n=", n_tested, ")")),
    vjust = -0.5, size = 3.2, colour = "grey20"
  ) +

  # Reference line at 50% — standard threshold for "high" resistance
  geom_hline(
    yintercept = 50, linetype = "dashed",
    colour = "#c0392b", linewidth = 0.8
  ) +
  annotate(
    "text", x = 5.4, y = 52,
    label = "50% threshold", hjust = 1, size = 3, colour = "#c0392b"
  ) +

  # Colour fill: red for high, amber for moderate, green for low
  scale_fill_manual(
    values = c("High (≥70%)"       = "#e74c3c",
               "Moderate (40–69%)" = "#e99e22",
               "Low (<40%)"        = "#37ae69"),
    name = "Resistance level"
  ) +

  # y-axis: 0–110% to leave room for labels above bars
  scale_y_continuous(
    limits = c(0, 110),
    labels = function(x) paste0(x, "%")
  ) +

  # Axis labels and title
  labs(
    title    = "Ciprofloxacin resistance by organism — 2022",
    subtitle = "First isolates only | Error bars = 95% Wilson Score CI | Dashed line = 50% threshold",
    x        = NULL,
    y        = "Resistance rate (%)",
    caption  = "Data source: Teaching dataset | EUCAST 2025 breakpoints"
  ) +

  # Apply our consistent theme
  theme(
    axis.text.x = element_text(face = "italic", size = 11)
  )

# Display the plot
p_bar
```

```{r bar-save, eval=FALSE}
# ── Save the plot as a high-quality PNG ──────────────────────────────────
# width and height are in inches; dpi = dots per inch (300 = print quality)
ggsave(
  filename = "bar_cip_resistance.png",
  plot     = p_bar,
  width    = 9,
  height   = 5.5,
  dpi      = 300
)
```

::: {.learn-box}
**Interpreting this chart**

Looking at the bar chart:

- **A. baumannii and P. aeruginosa** show very high ciprofloxacin resistance (red bars).
  This means standard fluoroquinolones cannot be used empirically for infections
  caused by these organisms at this hospital.
- **S. aureus** is in the moderate range — ciprofloxacin remains somewhat useful
  but should be confirmed by culture sensitivity.
- **E. coli and K. pneumoniae** also show concerning levels of fluoroquinolone resistance.

**Action implication:** The infection control team should review fluoroquinolone prescribing
and consider restricting empirical use for Gram-negative hospital-acquired infections.
:::

::: {.exercise-box}
**Exercise 1**

Modify the bar chart code above to show **meropenem resistance** instead of ciprofloxacin.


What do you notice about the meropenem resistance rates compared to ciprofloxacin?
Why is this clinically important?
:::

---

# Visualization 2: Histogram — MIC Distributions

::: {.learn-box}
**When to use a histogram in AMR surveillance**

A histogram shows you the **distribution** (spread and shape) of a set of numbers.
In AMR work, we use histograms for **MIC (minimum inhibitory concentration) values**
to understand how a population of bacteria responds to an antibiotic.

**Why this matters:**
- A unimodal (single-peak) distribution near the susceptible range is reassuring
- A bimodal (two-peak) distribution suggests a resistant subpopulation is emerging
- The position of the EUCAST/CLSI breakpoint relative to the peaks tells you
  how much of the population is captured on each side
:::

## Import MIC data

```{r mic-import}
# ── Import MIC values from the Excel workbook ─────────────────────────────
mic_data <- read_excel(excel_file, sheet = "MIC_Values")

# Standardise column names
mic_data <- mic_data %>% clean_names()

# Preview
head(mic_data, 6)
```

## Prepare MIC data for plotting

MIC values span a very wide range (0.03 to 64 mg/L or more).
We use a **log₂ scale** because each doubling of MIC is one step in the standard
serial dilution — so log₂ spacing is natural and standard in clinical microbiology.

```{r mic-prep}
# ── Focus on K. pneumoniae vs meropenem ──────────────────────────────────
kp_mem_mic <- mic_data %>%
  filter(
    organism == "K. pneumoniae",
    antibiotic == "Meropenem"
  ) %>%
  mutate(
    # Convert MIC to log2 scale
    mic_log2 = log2(mic_mg_l),
    # Clean up the interpretation column
    interpretation = str_to_upper(str_trim(interpretation)),
    interpretation = factor(interpretation, levels = c("S","I","R"))
  )

cat("Number of K. pneumoniae meropenem isolates:", nrow(kp_mem_mic), "\n")
cat("\nMIC value range:\n")
print(summary(kp_mem_mic$mic_mgl))
```

## Create the histogram

```{r hist-plot, fig.width=9, fig.height=5.5}
# ── EUCAST 2025 breakpoint for K. pneumoniae meropenem ───────────────────
# Susceptible: MIC ≤ 2 mg/L → log2(2) = 1
bp_log2 <- log2(2)

# ── Create x-axis labels (convert log2 back to mg/L) ─────────────────────
# Standard MIC steps in clinical microbiology
mic_steps    <- c(0.03, 0.06, 0.12, 0.25, 0.5, 1, 2, 4, 8, 16, 32, 64)
log2_steps   <- log2(mic_steps)
mic_labels   <- c("0.03","0.06","0.12","0.25","0.5","1","2","4","8","16","32","64")

# ── Build the histogram ───────────────────────────────────────────────────
p_hist <- ggplot(
  kp_mem_mic,
  aes(x = mic_log2, fill = interpretation)
) +

  # Histogram bars — binwidth = 1 means one bar per doubling dilution
  geom_histogram(
    binwidth = 1,
    colour   = "white",
    boundary = 0.5     # align bins to the standard MIC steps
  ) +

  # Breakpoint line
  geom_vline(
    xintercept = bp_log2 + 0.5,   # +0.5 puts line at the boundary between S and R
    linetype   = "dashed",
    colour     = "#2c3e50",
    linewidth  = 1.2
  ) +

  # Annotation explaining the breakpoint line
  annotate(
    "text",
    x = bp_log2 + 0.7, y = Inf,
    label  = "EUCAST breakpoint\n(S ≤2 mg/L, R >2 mg/L)",
    hjust  = 0, vjust = 1.5,
    size   = 3.5, colour = "#2c3e50"
  ) +

  # Shaded background to show S and R zones
  annotate("rect", xmin = -Inf, xmax = bp_log2 + 0.5,
           ymin = -Inf, ymax = Inf,
           fill = "#27ae60", alpha = 0.04) +
  annotate("rect", xmin = bp_log2 + 0.5, xmax = Inf,
           ymin = -Inf, ymax = Inf,
           fill = "#e74c3c", alpha = 0.04) +

  # Colour fill for S/I/R
  scale_fill_manual(
    values = c("S" = "#27ae60", "I" = "#e67e22", "R" = "#e74c3c"),
    name   = "Interpretation",
    labels = c("S" = "Susceptible (S)",
               "I" = "Intermediate (I)",
               "R" = "Resistant (R)")
  ) +

  # x-axis: show real MIC values, not log2 numbers
  scale_x_continuous(
    breaks = log2_steps,
    labels = mic_labels
  ) +

  labs(
    title    = "K. pneumoniae meropenem MIC distribution",
    subtitle = "Each bar = one standard doubling dilution | Dashed line = EUCAST 2025 breakpoint",
    x        = "MIC (mg/L) — log₂ scale",
    y        = "Number of isolates",
    caption  = "Data source: Teaching dataset | n = 40 isolates"
  )

p_hist
```

::: {.learn-box}
**Interpreting this histogram**

Look at the shape of the distribution:

- **Susceptible cluster** (left of the breakpoint): Most isolates have MICs of 0.06–1 mg/L,
  far below the EUCAST breakpoint of 2 mg/L. This suggests wild-type susceptibility.

- **Resistant cluster** (right of the breakpoint): A smaller group of isolates has MICs
  of 8–32 mg/L. This bimodal (two-peak) shape is a warning sign — it means a
  resistant subpopulation is present alongside the susceptible wild type.

- **Clinical implication:** Even though overall carbapenem resistance is moderate (~30%),
  the presence of a resistant cluster should trigger: (1) molecular testing for
  carbapenemase production (KPC, NDM, OXA), (2) strict contact precautions for
  carbapenem-resistant isolates, and (3) antibiotic stewardship review.
:::

::: {.exercise-box}
**Exercise 2**

The `MIC_Values` sheet also contains data for:

- `E. coli` vs `Ciprofloxacin`
- `A. baumannii` vs `Meropenem`

Repeat the histogram for one of these combinations.

Do you see a bimodal or unimodal distribution? What does the shape tell you?
:::

---

# Visualization 3: Box Plot — Comparing Distributions Across Groups

::: {.learn-box}
**When to use a box plot in AMR surveillance**

Box plots are useful when you want to compare the **spread and central tendency**
of MIC values across multiple groups at once — such as comparing MIC distributions
across different wards, different time periods, or different specimen types.

A single histogram shows you one group. A box plot shows you many groups side by side.

**Reading a box plot:**

- The **line inside the box** = median (middle value)
- The **box edges** = Q1 (25th percentile) and Q3 (75th percentile)
- The **whiskers** extend to 1.5× the interquartile range
- **Dots beyond the whiskers** = outliers (unusually high or low values)
:::

## Prepare data: MIC by ward

```{r box-prep}
# ── Filter to K. pneumoniae ciprofloxacin for box plot ───────────────────
kp_cip_mic <- mic_data %>%
  filter(
    organism   == "K. pneumoniae",
    antibiotic == "Ciprofloxacin"
  ) %>%
  mutate(
    mic_log2 = log2(mic_mg_l),
    ward     = factor(ward,
                      levels = c("ICU","SUR","MED","PED","OBS"))
  )

# How many isolates per ward?
kp_cip_mic %>%
  count(ward, name = "n") %>%
  kable(caption = "Table 3. K. pneumoniae isolates by ward") %>%
  kable_styling(bootstrap_options = "striped", full_width = FALSE)
```

## Create the box plot

```{r box-plot, fig.width=9, fig.height=5.5}
# EUCAST CIP breakpoint for K. pneumoniae: S ≤1, R >1
bp_cip_log2 <- log2(1)

p_box <- ggplot(
  kp_cip_mic,
  aes(x    = ward,
      y    = mic_log2,
      fill = ward)
) +

  # The box plots
  geom_boxplot(
    width         = 0.5,
    colour        = "grey30",
    outlier.shape = 21,          # outlier points as open circles
    outlier.size  = 2.5,
    outlier.fill  = "white",
    outlier.colour= "grey40"
  ) +

  # Overlay individual data points (jittered so they don't overlap)
  geom_jitter(
    width  = 0.05,
    alpha  = 0.5,
    size   = 1.8,
    colour = "grey20"
  ) +

  # Breakpoint reference line
  geom_hline(
    yintercept = bp_cip_log2,
    linetype   = "dashed",
    colour     = "#c0392b",
    linewidth  = 0.9
  ) +
  annotate(
    "text", x = 5.6, y = bp_cip_log2 + 0.15,
    label = "EUCAST\nbreakpoint\n(1 mg/L)",
    hjust = 1, size = 3, colour = "#c0392b"
  ) +

  # Ward colours: one colour per ward
  scale_fill_manual(
    values = c("ICU" = "#3498db", "SUR" = "#e67e22",
               "MED" = "#27ae60", "PED" = "#9b59b6",
               "OBS" = "#e74c3c"),
    guide  = "none"   # legend not needed — wards labelled on x-axis
  ) +

  # y-axis: show real MIC values
  scale_y_continuous(
    breaks = log2(c(0.06, 0.12, 0.25, 0.5, 1, 2, 4, 8, 16, 32)),
    labels = c("0.06","0.12","0.25","0.5","1","2","4","8","16","32")
  ) +

  labs(
    title    = "K. pneumoniae ciprofloxacin MIC distribution by ward",
    subtitle = "Each point = one isolate | Dashed line = EUCAST 2025 breakpoint (1 mg/L)",
    x        = "Ward",
    y        = "Ciprofloxacin MIC (mg/L) — log₂ scale",
    caption  = "Data source: Teaching dataset | First isolates only"
  )

p_box
```

::: {.learn-box}
**Interpreting this box plot**

- **MED** has the highest median MIC (box sits highest on the y-axis), meaning ICU
  isolates tend to be more resistant. This is expected — ICU patients receive more
  antibiotic courses, creating selection pressure.
- **Obstetrics** has the lowest median MIC — most isolates are susceptible.
- Any ward where the **median line is above the dashed breakpoint** has >50% of
  its isolates classified as resistant. This is a red flag.
- **Outlier dots** far above the main box indicate individual isolates with very high
  MICs — these could represent carbapenem-resistant or XDR strains worth investigating.

**Action:** Wards with high median MICs should have targeted stewardship reviews
and tighter infection control measures.
:::

---

# Visualization 4: Scatter Plot — Trends Over Time and Correlations

Scatter plots in AMR surveillance are used for two distinct purposes:

1. **Trend over time** — showing how resistance % changes across quarterly periods
2. **Correlation** — showing whether antibiotic consumption is associated with rising resistance

## 4.1 Import trend and consumption data

```{r trend-import}
# ── Import pre-aggregated quarterly resistance data ───────────────────────
quarterly_r <- read_excel(excel_file, sheet = "Quarterly_R") %>%
  clean_names()

# ── Import consumption data ───────────────────────────────────────────────
consumption <- read_excel(excel_file, sheet = "Consumption") %>%
  clean_names()

head(quarterly_r)
```

## 4.2 Trend scatter plot: Resistance over time

```{r scatter-trend, fig.width=9, fig.height=5}
# ── Prepare the trend data ────────────────────────────────────────────────
# We will plot K. pneumoniae ciprofloxacin resistance (kp_cip) over time

kp_trend <- quarterly_r %>%
  dplyr:: select(quarter, time_id, pct_r = kp_cip) %>%
  mutate(
    quarter = factor(quarter, levels = c("Q1-2022","Q2-2022",
                                         "Q3-2022","Q4-2022"))
  )

# ── Fit a linear regression model ────────────────────────────────────────
model <- lm(pct_r ~ time_id, data = kp_trend)
summary(model)

# Extract key statistics for the annotation
slope   <- round(coef(model)[2], 2)
r2      <- round(summary(model)$r.squared, 3)
p_value <- round(coef(summary(model))[2, 4], 4)

# Create prediction data for the regression line
pred_df <- data.frame(time_id = seq(1, 4, length.out = 50))
preds   <- predict(model, newdata = pred_df, interval = "confidence")
pred_df <- cbind(pred_df, preds)

# ── Build the scatter plot ────────────────────────────────────────────────
p_scatter <- ggplot(kp_trend, aes(x = time_id, y = pct_r)) +

  # Confidence interval ribbon (shaded area around regression line)
  geom_ribbon(
    data = pred_df,
    aes(x = time_id, ymin = lwr, ymax = upr),
    fill  = "#3498db", alpha = 0.12,
    inherit.aes = FALSE
  ) +

  # Regression line
  geom_line(
    data = pred_df,
    aes(x = time_id, y = fit),
    colour    = "#3498db",
    linetype  = "dashed",
    linewidth = 1,
    inherit.aes = FALSE
  ) +

  # Connecting line between points
  geom_line(colour = "grey60", linewidth = 0.7) +

  # Data points — coloured by resistance level
  geom_point(
    aes(fill = pct_r),
    shape = 21, size = 5,
    colour = "white", stroke = 1.2
  ) +

  # Colour scale: green → amber → red
  scale_fill_gradientn(
    colours = c("#27ae60", "#f39c12", "#e74c3c"),
    values  = rescale(c(0, 50, 100)),
    limits  = c(0, 100),
    name    = "Resistance (%)"
  ) +

  # Quarter labels on x-axis
  scale_x_continuous(
    breaks = 1:4,
    labels = levels(kp_trend$quarter)
  ) +

  scale_y_continuous(
    limits = c(0, 100),
    labels = function(x) paste0(x, "%")
  ) +

  # Regression equation annotation
  annotate(
    "text",
    x = 1, y = 95,
    label = paste0(
      "β = +", slope, "% per quarter\n",
      "R² = ", r2, "\n",
      "p = ", p_value
    ),
    hjust  = 0, vjust = 1,
    size   = 3.5, colour = "#2c3e50",
    fontface = "italic"
  ) +

  labs(
    title    = "K. pneumoniae ciprofloxacin resistance — quarterly trend",
    subtitle = "Dashed line = linear regression | Shaded area = 95% confidence interval",
    x        = "Quarter",
    y        = "Resistance rate (%)",
    caption  = "Data source: Teaching dataset | First isolates only"
  )

p_scatter
```

::: {.learn-box}
**Interpreting the trend plot**

- **Beta (β):** The slope of the regression line, in units of "% change per quarter."
  A positive β (e.g., +3.2%) means resistance is **rising** each quarter.
- **R²:** How well the time variable explains the variation in resistance.
  R² = 0.95 means 95% of the variation in resistance is explained by time — a strong trend.
- **p-value:** Tests whether the slope is statistically significantly different from zero.
  p < 0.05 means the trend is unlikely to be due to chance alone.

**Important limitation:** With only 4 time points, the statistical power is low.
Even a significant p-value should be interpreted cautiously with so few data points.
Always note this limitation when reporting.
:::

## 4.3 Correlation: Consumption vs Resistance

```{r scatter-corr, fig.width=8, fig.height=5.5}
# ── Summarise ciprofloxacin consumption by quarter ────────────────────────
cip_consumption <- consumption %>%
  filter(str_detect(drug_name, "Cipro|cipro")) %>%
  group_by(quarter) %>%
  summarise(
    cip_ddd_1000pd = round(sum(ddd_dispensed) / sum(patient_days) * 1000, 1),
    .groups = "drop"
  )

# Join with resistance data
corr_df <- left_join(
  cip_consumption,
  quarterly_r %>% dplyr::select(quarter, kp_cip),
  by = "quarter"
) %>%
  filter(!is.na(kp_cip))

# Pearson correlation
cor_test <- cor.test(corr_df$cip_ddd_1000pd, corr_df$kp_cip, method = "pearson")
cat(sprintf("Pearson r = %.3f | p = %.4f\n",
            cor_test$estimate, cor_test$p.value))

# ── Scatter: DDD vs resistance ────────────────────────────────────────────
p_corr <- ggplot(corr_df,
                 aes(x = cip_ddd_1000pd, y = kp_cip,
                     label = quarter)) +

  geom_smooth(method = "lm", se = TRUE,
              colour = "grey60", fill = "grey90",
              linewidth = 0.8) +

  geom_point(aes(colour = quarter), size = 5, shape = 17) +

  geom_text(vjust = -1, hjust = 0.5, size = 3.5, fontface = "bold") +

  scale_colour_manual(
    values = c("Q1-2022" = "#3498db", "Q2-2022" = "#e67e22",
               "Q3-2022" = "#e74c3c", "Q4-2022" = "#27ae60"),
    guide = "none"
  ) +

  scale_y_continuous(limits = c(65, 95),
                     labels = function(x) paste0(x, "%")) +

  labs(
    title    = "Ciprofloxacin consumption vs K. pneumoniae resistance",
    subtitle = paste0("Pearson r = ", round(cor_test$estimate, 3),
                      " | Ecological correlation — not individual-level causation"),
    x        = "Ciprofloxacin DDD per 1,000 patient days",
    y        = "K. pneumoniae CIP resistance (%)",
    caption  = "Each point = one quarter | Regression line with 95% CI shown"
  )

p_corr
```

::: {.danger-box}
**The Ecological Fallacy — a critical warning**

When we correlate ward-level antibiotic consumption with ward-level resistance rates,
we are doing an **ecological study**. This means:

- We cannot say that *patients who received ciprofloxacin* are more likely to become resistant
- Many other factors (clonal outbreak, patient mix, infection control) drive resistance
- The correlation may be spurious, especially with only 4 quarterly data points

Always add this caveat when reporting consumption-resistance correlations.
:::

---

# Visualization 5: Heatmap — The Full Resistance Overview

::: {.learn-box}
**When to use a heatmap in AMR surveillance**

A heatmap is the **information-densest** AMR visualization. It shows resistance rates
for many organism-antibiotic combinations at once using colour intensity.

It is the standard format for:
- Annual antibiogram presentations to infection control committees
- Hospital-level AMR reports to management
- Quick identification of which organism-antibiotic combinations are in crisis
:::

## Calculate all resistance rates

```{r heatmap-calc}
# ── Calculate resistance for all priority organism-antibiotic pairs ────────
# We build a helper function then call it for each pair

pairs <- list(
  list(org="Klebsiella pneumoniae",   ab="cip"), list(org="Klebsiella pneumoniae",   ab="mem"),
  list(org="Klebsiella pneumoniae",   ab="ctx"), list(org="Klebsiella pneumoniae",   ab="gen"),
  list(org="Escherichia coli",        ab="amp"), list(org="Escherichia coli",        ab="ctx"),
  list(org="Escherichia coli",        ab="cip"), list(org="Escherichia coli",        ab="sxt"),
  list(org="Staphylococcus aureus",   ab="oxa"), list(org="Staphylococcus aureus",   ab="gen"),
  list(org="Staphylococcus aureus",   ab="cip"),
  list(org="Pseudomonas aeruginosa",  ab="mem"), list(org="Pseudomonas aeruginosa",  ab="cip"),
  list(org="Acinetobacter baumannii", ab="mem"), list(org="Acinetobacter baumannii", ab="cip"),
  list(org="Acinetobacter baumannii", ab="col")
)

all_resist <- bind_rows(lapply(pairs, function(p) {
  calc_resistance_ci(clean_data, p$org, p$ab)
}))

# Short organism names
all_resist <- all_resist %>%
  mutate(
    org_short = case_when(
      organism == "Klebsiella pneumoniae"   ~ "K. pneumoniae",
      organism == "Escherichia coli"        ~ "E. coli",
      organism == "Staphylococcus aureus"   ~ "S. aureus",
      organism == "Pseudomonas aeruginosa"  ~ "P. aeruginosa",
      organism == "Acinetobacter baumannii" ~ "A. baumannii",
      TRUE ~ organism
    ),
    # Flag small sample sizes
    label = paste0(pct_R, "%", ifelse(n_tested < 30, "*", ""))
  )

# Preview
all_resist %>%
  dplyr::select(org_short, antibiotic, n_tested, pct_R, label) %>%
  kable(caption = "Table 4. Resistance rates for heatmap") %>%
  kable_styling(bootstrap_options = "striped")
```

## Create the heatmap

```{r heatmap-plot, fig.width=11, fig.height=5.5}
# ── Order organisms by clinical priority ──────────────────────────────────
org_order <- c("A. baumannii","P. aeruginosa","K. pneumoniae",
                "E. coli","S. aureus")
ab_order  <- c("COL","MEM","CTX","AMP","CIP","GEN","SXT","OXA")

all_resist <- all_resist %>%
  mutate(
    org_short  = factor(org_short,  levels = org_order),
    antibiotic = factor(antibiotic, levels = ab_order)
  )

p_heat <- ggplot(
  all_resist,
  aes(x = antibiotic, y = org_short, fill = pct_R)
) +

  # Tiles — white lines between tiles act as grid lines
  geom_tile(colour = "white", linewidth = 1.5) +

  # Text label inside each tile
  geom_text(
    aes(label = label,
        colour = pct_R > 50),   # white text on dark tiles, dark on light
    size      = 3.5,
    fontface  = "bold"
  ) +

  # Colour scale: green → amber → red
  scale_fill_gradientn(
    colours   = c("#27ae60", "#f1c40f", "#e74c3c"),
    values    = rescale(c(0, 40, 100)),
    limits    = c(0, 100),
    na.value  = "grey85",
    name      = "Resistance (%)",
    guide     = guide_colourbar(barwidth = 12, barheight = 0.5,
                                title.position = "top",
                                title.hjust    = 0.5)
  ) +

  # Text colour: white on high-resistance tiles, dark on low
  scale_colour_manual(values = c("TRUE" = "white", "FALSE" = "#2c3e50"),
                      guide = "none") +

  labs(
    title    = "Antimicrobial resistance heatmap — 2022",
    subtitle = "Priority pathogens × Antibiotics | *n < 30 isolates (interpret with caution) | Grey = not tested",
    x        = "Antibiotic",
    y        = NULL,
    caption  = "Data source: Teaching dataset | First isolates only | EUCAST 2025 breakpoints"
  ) +

  theme(
    axis.text.y     = element_text(face = "italic", size = 11),
    axis.text.x     = element_text(face = "bold", size = 11),
    legend.position = "bottom",
    panel.grid      = element_blank()
  )

p_heat
```

::: {.learn-box}
**Reading the heatmap**

- **Dark red tiles** = high resistance — these organism-antibiotic combinations
  are failing and cannot be used empirically
- **Green tiles** = low resistance — these remain effective
- **Grey tiles** = not tested (e.g., colistin not tested for S. aureus because
  Gram-positives are intrinsically resistant)
- **Asterisk (*)** = fewer than 30 isolates — the rate is unreliable

**Pattern to look for:**
If an entire row is red (like A. baumannii), that organism is approaching XDR
(extensively drug resistant) status and management requires specialist input.
:::

---

# Combining Charts: Creating a Dashboard Figure

For reports and presentations, you often want to show multiple related charts together.
The `patchwork` package makes this easy:

```{r dashboard, fig.width=13, fig.height=10}
# ── Combine the bar plot and heatmap into one figure ──────────────────────
combined <- (p_bar / p_heat) +
  plot_annotation(
    title   = "AMR Surveillance Dashboard — Kpandai District Hospital, 2022",
    caption = "First isolates only | EUCAST 2025 breakpoints | *n < 30",
    theme   = theme(
      plot.title   = element_text(size = 16, face = "bold",
                                   colour = "#1A4A8A", hjust = 0.5),
      plot.caption = element_text(size = 9, colour = "grey50")
    )
  ) +
  plot_layout(heights = c(1.2, 1))

combined
```

```{r dashboard-save, eval=FALSE}
# ── Save dashboard figure ─────────────────────────────────────────────────
ggsave(
  "AMR_Dashboard_2022.png",
  plot   = combined,
  width  = 13,
  height = 10,
  dpi    = 300
)
```

---

# Putting It All Together: A Complete Reporting Workflow

```{r full-workflow, eval=FALSE}
# ── COMPLETE WORKFLOW — COPY THIS FOR YOUR OWN HOSPITAL DATA ──────────────

# 1. Set your file path
my_file <- "path/to/your/data.xlsx"

# 2. Import
raw <- read_excel(my_file, sheet = "YourSheetName") %>% clean_names()

# 3. Clean
clean <- raw %>%
  filter(!is.na(patient_id)) %>%                          # remove blanks
  mutate(organism_std = str_to_title(str_trim(organism))) # standardise names

# 4. First isolate rule
analysis <- clean %>%
  arrange(patient_id, organism_std, collection_date) %>%
  group_by(patient_id, organism_std) %>%
  slice(1) %>%  # keep only the first row per group
  ungroup()

# 5. Calculate resistance
results <- calc_resistance_ci(analysis, "Klebsiella pneumoniae", "cip")

# 6. Plot
p <- ggplot(results, aes(x = antibiotic, y = pct_R, fill = pct_R)) +
  geom_col() + amr_theme

# 7. Save
ggsave("my_resistance_plot.png", plot = p, width = 8, height = 5, dpi = 300)
```

---

# Session Information

It is good practice to include session information at the end of your R Markdown document.
This records exactly which versions of R and each package were used — important for
reproducibility.

```{r session-info}
sessionInfo()
```

---

::: {.tip-box}
**Next steps **

1. **Exercise :** Use this Rmd with your own hospital's WHONET export. Follow the same cleaning steps.

2. Explore the `AMR` package in R — purpose-built for clinical microbiology data analysis with built-in EUCAST/CLSI breakpoints.
:::

---

*AMR Surveillance Training Programme |  Data Visualization in R | Version 1.0, 2026*
